---
title: "Poverty Metrics Analysis in DRC"
author: "Elvira Khwatenge"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(xfun)
# Load the dataset
data <- read.csv("hdx_hapi_poverty_rate_global.csv")

```

## Introduction
In this report, we explore global poverty metrics, focusing on indicators such as the Multidimensional Poverty Index (MPI), headcount ratio, intensity of deprivation, vulnerability to poverty, and severe poverty levels using data cleaning, descriptive statistics, and visualization techniques to reveal trends, regional disparities, and correlations.

```{r}
# Clean data
colnames(data) <- data[1, ]  # Use the first row as column headers
data <- data[-1, ]  # Remove the first row now
colnames(data) <- c(
  "location_code", "has_hrp", "in_gho", "provider_admin1_name", 
  "admin1_code", "admin1_name", "mpi", "headcount_ratio", 
  "intensity_of_deprivation", "vulnerable_to_poverty", 
  "in_severe_poverty", "reference_period_start", "reference_period_end"
)
numeric_cols <- c("mpi", "headcount_ratio", "intensity_of_deprivation", 
                  "vulnerable_to_poverty", "in_severe_poverty")
data[numeric_cols] <- lapply(data[numeric_cols], as.numeric)

# Display structure
str(data)

# Summary statistics
summary(data[numeric_cols])
```


```{r}
library(ggplot2)

# Histogram of MPI
ggplot(data, aes(x = mpi)) +
  geom_histogram(binwidth = 0.05, fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of MPI", x = "MPI", y = "Frequency")
```

## Trends


```{r}
# Convert dates to Date type
data$reference_period_start <- as.Date(data$reference_period_start)
data$reference_period_end <- as.Date(data$reference_period_end)

# Line plot of MPI over time
ggplot(data, aes(x = reference_period_start, y = mpi, color = admin1_name)) +
  geom_line() +
  theme_minimal() +
  labs(title = "MPI Trends Over Time", x = "Year", y = "MPI")


```


## MPI vs. Headcount Ratio

```{r}
# Scatter plot of MPI vs. Headcount Ratio
ggplot(data, aes(x = mpi, y = headcount_ratio)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "MPI vs Headcount Ratio", x = "MPI", y = "Headcount Ratio")
```


## Regional Comparison
Compare poverty indicators (e.g., MPI, headcount ratio) across different regions (admin1_name).

# Average MPI by Region
```{r}
# Average MPI by region
library(dplyr)

regional_mpi <- data %>%
  group_by(admin1_name) %>%
  summarize(avg_mpi = mean(mpi, na.rm = TRUE)) %>%
  arrange(desc(avg_mpi))

# Bar plot
ggplot(regional_mpi, aes(x = reorder(admin1_name, -avg_mpi), y = avg_mpi)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Average MPI by Region", x = "Region", y = "Average MPI") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Poverty Trends Over Time
Analyze how poverty indicators (e.g., mpi, headcount_ratio) evolve over time.

# Headcount Ratio Over Time
```{r}

# Line plot of headcount ratio over time
ggplot(data, aes(x = reference_period_start, y = headcount_ratio, color = admin1_name)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Headcount Ratio Trends Over Time", x = "Year", y = "Headcount Ratio")
```


## Correlation Between Indicators
Explore the relationship between poverty indicators, such as mpi, headcount_ratio, and intensity_of_deprivation.

# Correlation Matrix
```{r}

# Correlation matrix of numeric variables
library(corrplot)

numeric_data <- data[numeric_cols]
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Correlation plot
corrplot(cor_matrix, method = "circle", type = "upper", tl.cex = 0.8)
```


## Severity Analysis
Identify regions or time periods with severe poverty levels.

Highlight Regions with Severe Poverty
```{r}
# Filter data for severe poverty
severe_poverty <- data %>%
  filter(in_severe_poverty > 50)  # Example threshold for severe poverty

# Table of regions with severe poverty
severe_poverty %>%
  select(admin1_name, reference_period_start, in_severe_poverty) %>%
  arrange(desc(in_severe_poverty))
```

## Vulnerable Populations
Analyze the percentage of people vulnerable to poverty (vulnerable_to_poverty).
## Vulnerability to Poverty by Region
```{r}

# Box plot for vulnerable populations by region
ggplot(data, aes(x = admin1_name, y = vulnerable_to_poverty)) +
  geom_boxplot(fill = "lightgreen", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Vulnerability to Poverty by Region", x = "Region", y = "Vulnerability (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Outlier Detection
Detect extreme values for poverty metrics.

```{r}

# Boxplot for MPI to detect outliers
ggplot(data, aes(x = "", y = mpi)) +
  geom_boxplot(fill = "orange", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Outliers in MPI", x = "", y = "MPI")
```

## Combine Insights
Generate an integrated dashboard with multiple visualizations using the flexdashboard or shiny package.

```{r}

# Example: Shiny Dashboard
library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Poverty Indicators Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput("indicator", "Select Indicator", choices = numeric_cols)
    ),
    mainPanel(
      plotOutput("trendPlot")
    )
  )
)

server <- function(input, output) {
  output$trendPlot <- renderPlot({
    ggplot(data, aes(x = reference_period_start, y = !!sym(input$indicator), color = admin1_name)) +
      geom_line() +
      theme_minimal() +
      labs(title = paste(input$indicator, "Trends Over Time"), x = "Year", y = input$indicator)
  })
}

shinyApp(ui = ui, server = server)
```

